---
title: "MLB Run Projection - HW#2, Question 2"
author: "Anthony Stachowski, Brian Krupa, Rory Kelly"
date: "4/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in CSV file containing data on 2019 MLB Season
File is stored at Github

```{r, message=FALSE}

library(readr)

urlfile = "https://raw.githubusercontent.com/AS-72/Sports-Analytics/master/MLB_Team_Data_2019.csv"

              
mlb_2019_data <- read_csv(urlfile, col_names = TRUE)

mlb_2019_data$Tm = as.factor(mlb_2019_data$Tm)

head(mlb_2019_data)

```

## Run multiple liner regression model on table:
This model uses TB = (Singles + 2xDoubles + 3xTriples + 4xHomeruns), BB, HBP, and SB to project runs by each team.
The variables are a measure of the number of base runners.

```{r}

runs_fit = lm(R ~ TB + BB + HBP + SB, data = mlb_2019_data)
summary(runs_fit)

fitted(runs_fit)

residuals(runs_fit)
```

#### The Adjusted R-squared measure for this model will give an indication as to how it performs.
The adjusted R-squared value accounts for additional predictors that are added to the model as increasing variables will increase normal R-squared values.  In the above model, the adjusted R-squared value is 93.32%.  Thus the model is expected to explain 93.32% of variation in runs by the variables included.


#### Display of regression diagnostics:
Below are four diagrams to assess the above linear model.

```{r}

par(mfrow=c(2,2))
plot(runs_fit, which = 1:4)

```

The Residuals vs. Fitted plot is a check for homoscedasticity and patterns would be exhibitied if this assumption failed. The plot shows that the residuals vary randomly around 0 and show no strong patterns based on 30 observations.

The Normal Q-Q plot should track closely with the y=x line, there appear to be some issues in the tail, which may call the normality assumption into question. However, as only 30 observations are considered, there will be a bit more variability.

The Scale-Location chart is a check for homoscedasticity and we are looking to ensure there are no extreme patterns. There do not appear to be any in this model.

Cook's Distance chart shows that observation 10 is larger than other observations, but it still remains below 1.00. This observation may require further investigation as the model seems to do a poor job in projecting the final runs for the season.

**Based on these regression diagnostics and the adjusted R-squared value, the model does a relatively good job in projecting the total runs by each team for a season.**

#### p-Values for examining coefficients:
Three of the 5 coefficients (including the intercept) in the model are significant at p=0.05. These variables are: intercept, total bases (TB), and walks (BB). 

One of the coefficients, stolen bases (SB), was just outside of this range at p=0.0635 and is significant at p=0.10.

The remaining coefficient, hit-by-pitch (HBP), was not significant with a p-value of 0.3875 and can be removed from the model.
